The Efficient Handling of Guards \\in the Design of RPython's Tracing JIT

### 2012-08-17 talk/vmil2012/paper.tex

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations about guards and
describe based on them the reasoning behind their implementation in
RPython's tracing just-in-time compiler. The contributions of this paper are:
\begin{itemize}
  \item an analysis and benchmark of guards in the context of RPython's tracing JIT,
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of the JIT and describe the rationale behind the design
\end{itemize}

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe for RPython's tracing JIT the details of guards in
the frontend. In this context the frontend is concerned with recording and storing the
information required to rebuild the interpreter state in case of a guard
failure. Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section~\ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally, Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project~\cite{rigo_pypys_2006} were started
in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension.\footnote{\url{http://pypy.org}} PyPy is now a fully compatible
alternative interpreter for the Python language.
Using RPython's tracing JIT compiler it is on average about 5 times faster than
CPython, the reference implementation.
PyPy is an interpreter written in RPython and takes advantage of the language
features provided by RPython
such as the provided tracing just-in-time compiler described below.

@@

different levels of completeness.

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object-oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

@@

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

Since tracing linearizes control flow by following one concrete execution,
the full control flow of a program is not observed.
The possible points of deviation from the trace are denoted by guard operations
that check whether the same assumptions observed while tracing
still hold during execution.
Similarly, in the case of dynamic languages guards can also encode type
assumptions.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

@@

%\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
\begin{lstlisting}[xleftmargin=20pt,xrightmargin=20pt,framexleftmargin=5pt,framexrightmargin=-10pt,mathescape, numbers=right, escapechar=|, firstnumber=18,frame=b]
$b_3$ = int_eq($i_5$, 1)                    |\setcounter{lstnumber}{17}|
guard_false($b_3$)                          |\setcounter{lstnumber}{-1}|
\end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Result of separated (left) and merged (right) compilation of one guard and the following operation (top).}
  \label{fig:trace-compiled}
\end{figure}

@@

First, a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that requires
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard~\#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard~\#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

\subsection{Space Overhead of Guards}
\label{sub:guard_overhead}
\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume data sizes}
    \label{fig:resume_data_sizes}
\end{figure}

@@

@@

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently using the techniques described earlier. On the other hand
comparing the results to the xz compression which only needs between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good, compressed representation of the resume data for the
large amount of guards present in the traces.

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the compiled traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bridge is compiled for them, only 1.2\% to 3.6\% of all guards
fail often enough that a bridge is compiled. Also, of all failing guards a few fail extremely often
and most fail rarely. Reinforcing this notion the figure shows that, depending on the
benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
of the total guards failures.
These results emphasize that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions plans to switch to compressed snapshots to further reduce
redundancy. The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.

>>>>

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations about guards and
describe based on them the reasoning behind their implementation in
RPython's tracing just-in-time compiler. The contributions of this paper are:
\begin{itemize}
  \item An analysis guards in the context of RPython's tracing JIT,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level components of RPython's JIT and a description of the rationale behind the design.
\end{itemize}

The set of central concepts upon which this work is based are described in
Section~\ref{sec:Background}, such as the PyPy project, the RPython language
and its meta-tracing JIT. Based on these concepts in Section~\ref{sec:Resume
Data} we proceed to describe the details of guards in
the frontend of RPython's tracing JIT.
Once the frontend has traced and optimized a loop it invokes the
backend to compile the operations to machine code, Section~\ref{sec:Guards in
the Backend} describes the low-level aspects of how guards are implemented in
the machine specific JIT-backend. The frequency of guards and the overhead associated with the
implementation described in this paper is discussed in
Section~\ref{sec:evaluation}. Section~\ref{sec:Related Work} presents an
overview about how guards are treated in the context of other just-in-time
compilers. Finally, Section~\ref{sec:Conclusion} summarizes our conclusions and
gives an outlook on further research topics.

@@

The RPython language and the PyPy project\footnote{\url{http://pypy.org}}~\cite{rigo_pypys_2006} were started
in 2002 with the goal of
creating a Python interpreter written in a high level language, allowing easy
language experimentation and extension. PyPy is now a fully compatible
alternative interpreter for the Python language.
Using RPython's tracing JIT compiler it is on average about 5 times faster than
CPython, the reference implementation.
PyPy is an interpreter written in RPython and takes advantage of the language
features provided by RPython
such as the provided tracing just-in-time compiler described below.

@@

RPython can mean one of two things:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language, is a subset of Python that provides a
statically typed object-oriented high level language. The subset of Python available in RPython is chosen in a way type inference is possible\cite{ancona_rpython:_2007}.
The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the interpreter into a C program.\footnote{
    RPython can also be used to translate programs to CLR and Java
    bytecode~\cite{ancona_rpython:_2007}, but this feature is somewhat
    experimental.
}
During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as a garbage collector and a tracing JIT compiler.
The process of inserting a tracing JIT is not fully automatic but is guided by
hints from the interpreter author.

@@

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution, the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}. These operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing\cite{Gal:2009ux}.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

In this context we refer to frontend as the component of the JIT that is
concerned with recording and optimizing the traces as well as storing the
information required to rebuild the interpreter state in case of a guard
failure.
Since tracing linearizes control flow by following one concrete execution,
the full control flow of a program is not observed.
The possible points of deviation from the trace are denoted by guard operations
that check whether the same assumptions observed while tracing
still hold during execution.
Similarly, in the case of dynamic languages guards can also encode type
assumptions.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

@@

%\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
\begin{lstlisting}[xleftmargin=20pt,xrightmargin=20pt,framexleftmargin=5pt,framexrightmargin=-10pt,mathescape, numbers=right, escapechar=|, firstnumber=18,frame=b]
$b_3$ = int_eq($i_5$, 1)                    |\setcounter{lstnumber}{17}|
guard_false($b_3$)                          |\setcounter{lstnumber}{-1}|
\end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
...
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Result of separated (left) and merged (right) compilation of one guard and the following operation (top).}
  \label{fig:trace-compiled}
\end{figure}

@@

First, a special data
structure called \emph{backend map} is created. This data structure encodes the
mapping from IR-variables needed by the guard to rebuild the state to the
low-level locations (registers and stack) where the corresponding values will
be stored when the guard is executed.
This data
structure stores the values in a succinct manner. The encoding is efficient to create and
provides a compact representation of the needed information in order
to maintain an acceptable memory profile.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split.
The guard becomes the branching point of two conditional paths with no
additional overhead.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard~\#1 jumps to the trampoline, loads the backend map and
then calls the bailout handler, whereas Guard~\#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the compiled traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer, only 1.2\% to 3.6\% of all guards fail often
enough that a bridge is compiled for them.
Also, of all failing guards a few fail extremely often
and most fail rarely. Reinforcing this notion the figure shows that, depending on the
benchmark, between 0.008\% and 0.225\% of the guards are responsible for 50\%
of the total guards failures.
These results emphasize that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

\subsection{Space Overhead of Guards}
\label{sub:guard_overhead}

@@

\begin{figure}
    \include{figures/resume_data_table}
    \caption{Resume data sizes}
    \label{fig:resume_data_sizes}
\end{figure}

@@

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently using the techniques described earlier. On the other hand
comparing the results to the xz compression which only needs between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal and could be improved taking into account the trade-off between
the required space and the time needed to build a good, compressed
representation of the resume data for the large amount of guards present in the
traces.

@@

Mike Pall, the author of LuaJIT describes in a post to the lua-users mailing
list different technologies and techniques used in the implementation of
LuaJIT~\cite{Pall:2009}. Pall explains that guards in LuaJIT use a datastucture
called snapshots, similar to RPython's resume data, to store the information
about how to rebuild the state from a guard failure using the information in the
snapshot and the machine execution state. According to Pall~\cite{Pall:2009}
snapshots for guards in LuaJIT are associated with a large memory footprint.
The solution used there is to store sparse snapshots, avoiding the creation
of snapshots for every guard to reduce memory pressure. Snapshots are only
created for guards after updates to the global state, after control flow points
from the original program and for guards that are likely to fail. As an outlook
Pall mentions plans to switch to compressed snapshots to further reduce
redundancy.\footnote{This optimization is now implemented in LuaJIT, at the time of writing it has not been fully documented in the LuaJIT Wiki (\url{http://wiki.luajit.org/Optimizations\#1-D-Snapshot-Compression}).}
The approach of not creating snapshots at all for every guard is
orthogonal to the resume data compression presented in this paper and could be
reused within RPython to improve the memory usage further.
